---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

```{r, echo = F}
suppressPackageStartupMessages(library(ph2rand))
```

# ph2rand

*Design of randomized comparative phase II oncology trials*

## Description

__ph2rand__ provides a suite of functions to assist with the design and visualization of randomized comparative phase II oncology trials. Specifically, support is provided to perform a sample size calculation for each of the most popular randomized comparative phase II oncology trial designs, for both point and composite null hypotheses. A selection of plotting functions also permit the informative depiction of various important quantities.

## Getting started

You can install the the latest development version of __ph2rand__, available from [GitHub](https://github.com/), with:

```{r, eval = F}
devtools::install_github("mjg211/ph2rand")
```

An introductory example of how to make use of the package’s core functionality can be found below. More detailed support is available in the package vignette, which can be accessed with `vignette("ph2rand")`. For further help, please contact Michael Grayling at michael.grayling@newcastle.ac.uk.

## Example

...